---
title: "practical - 7 :  Principal component Analysis and PCA in regression"
author: "SHRUTI KHARVA"
date: "2026-03-25"
output: html_document
---

Q.1	A study was conducted to analyze the relationship among seven variables. The covariance matrix of the variables X_1,X_2,…,X_7is given below:

S=    ■(1.307&0.879&0.908&0.958&0.345&0.512&0.385@&1.248&1.147&1.411&-0.089&0.564&0.172@&&1.317&1.480&0.058&0.573&0.259@&&&2.022&-0.110&0.773&0.236@&&&&0.443&0.046&0.259@&&&&&0.371&0.154@&&&&&&0.264)

	Determine the principal components (PCs).
	Interpret the first few principal components.


```{r}
#Find principal component when S = cov(x) matrix is given
S = matrix(c(1.307,0.879,0.908,0.958,0.345,0.512,0.385,0,1.248,1.147,1.411,-0.089,0.564,0.172,0,0,1.317,1.480,0.058,0.573,0.259,0,0,0,2.022,-0.110,0.773,0.236,0,0,0,0,0.443,0.046,0.259,0,0,0,0,0,0.371,0.154,0,0,0,0,0,0,0.264),byrow=TRUE,7)
S
s = S+t(S)-diag(diag(S))
s
p = 7 #number of indepedent variables
n = 36 #We have 36 observation
l = eigen(s)$values
l
g = eigen(s)$vectors
g #eigen vectors are orthogonal to each other
g=round(g,4)

propvar = (l/sum(l))*100 # proportion variance explained by ith component
propvar
cumprop=cumsum(propvar) # proportion of variance explained by frist i component
cumprop

#the eigenvalues and the percentage of variance explained by the component
Pc_no = 1:length(l)
cbind(Pc_no,l,propvar,cumprop)

#conclusion : It is since that the frist component explains by 76% variation dataset explains by X1 followed by X2,followed by X3 and followed by X4
#The second component explains 15% variation of data explains by X2 followed by The frist three component explaines around 95% variation of the dataset. 

#calculate loading
rij = (t(g)*sqrt(l))/sqrt(diag(s))
rij = t(rij);rij
rij^2 #Weights

# LRT to select the number of components
for (r in 0:(p-1))
{
  a0 = sum(l[(r+1):p])/(p-r)
  g0 = prod(l[(r+1):p])^(1/(p-r))
  T = (n - ((2*p + 11)/6)) * (p - r) * log(a0/g0)
  df = (p - r + 2) * (p - r - 1) / 2
  pvalue = 1 - pchisq(T, df)
  if (pvalue > 0.05) { break }
}

cat("Here accept H0 for r=", r, "chisq=", T, "p-value=", pvalue)

# scree plot
plot(1:7, l, type="o", col="darkblue")


```
Q.2	A study was conducted on 60 children (n = 60) to analyze the relationship between a response variable and several explanatory variables related to child characteristics. The variables recorded are:
y : Response variable 
x₁, x₂, x₃, x₄, x₅, x₆, x₇, x₈ : Predictor variables representing different child-related factors
i.	Obtain the correlation matrix of the predictor variables.
ii.	Determine the principal components (PCs).
iii.	Interpret the first few principal components.
iv.	Fit the Principal Component Regression model.
v.	Express the regression equation: In terms of principal components and In terms of original variables

USE CHILD DATASET.

```{r}
#P
```

