---
title: "Test Concerning Mean Vector (Two Sample Problem)"
author: "Alokchandra Jha"
date: "2026-01-01"
output: html_document
---
1. 	Using Moody’s bond rating data sample of 20 “Aa” (middle high quality) corporate bonds and 20 “Baa”( Top medium quality ) corporate bonds were selected. For each of the corresponding companies, the ratios 
X1 : Current ratio ( measure of short term liquidity)
X2 : Time long-term interest rate (measure of financial risk as interest coverage) 
X3 : debt to equity ratio ( a measure of financial risk as leverage)
X4 : rate of return on equity ( A measure  of profitability)
Were recorded the summary statistics are as follows:
N1 = 20   N2 = 20 
Are financial characteristics of firm with Ab bond different from those with Baa bonds at 
α = 0.05
Aa bond companies :                                             
X ̅ 1 =  ■(2.287&12.600&0.34477&14.830)
S1 = ■(0.459&0.254&-0.026&-0.244@0.254&27.465&-0.589&-0.267@-0.026&-0.589&0.030&0.102@-0.244&-0.267&0.102&6.854)
Baa bond companies:
X ̅ 2 =  ■(2.404&7.155&0.524&12.840)

S2 = ■(0.944&-0.089&0.002&-0.719@-0.089&16.432&-0.400&19.044@0.002&-0.400&0.024&-0.094@-0.719&19.044&-0.094&61.854)

```{r}
#equal sample size
N = 20

P = 4
#to test if financial characteristics of both the bonds is same or different

#Hypothesis :
#Ho : There is no difference between the means of Aa and Baa bonds. (X1_bar = X2_bar)
#H1 : There is a difference between the means of Aa and Baa bonds. (X1_bar != X2_bar)

x1_bar = c(2.287, 12.600, 0.34477, 14.830)
x2_bar = c(2.404, 7.155, 0.524, 12.840)

y_bar = x1_bar - x2_bar
y_bar

s1 = matrix(c(0.459, 0.254, -0.026, -0.244, 0.254, 27.465, -0.589, -0.267, -0.026, -0.589, 0.030, 0.102, -0.244, -0.267, 0.102, 6.854), nrow = 4, ncol = 4, byrow = TRUE)
s2 = matrix(c(0.944, -0.089, 0.002, -0.719, -0.089, 16.432, -0.400, 19.044, 0.002, -0.400, 0.024, -0.094, -0.719, 19.044, -0.094, 61.854), nrow = 4, ncol = 4, byrow = TRUE)
s1
s2

y_s = s1 + s2
y_s

hotelling_t = N %*% t(y_bar) %*% solve(y_s) %*% (y_bar)
hotelling_t

f1 = ((N-P)/ P) * hotelling_t/(N-1)
f1

f_tab = qf(1-0.05, P, N-P)
f_tab

if (f1 > f_tab) {
  print("We reject the null hypothesis at 5% level of Significance")
} else {
  print("We do not have enough evidence to reject the null hypothesis at 5% level of Significance")}
print("Hence, we conclude that there is a statistically significant difference among the financial characteristics of Aa and Baa bonds.")
```
```{r}
data1 = as.matrix(read.csv("C:\\Users\\statlab\\Desktop\\AlokPrev\\Multivariate\\Practical2\\M.A.-4IRIS SETOSA DATA.csv"))
data2 = as.matrix(read.csv("C:\\Users\\statlab\\Desktop\\AlokPrev\\Multivariate\\Practical2\\M.A.-4IRIS VERSICOLOR DATA.csv"))

number_rows = nrow(data1)

P = ncol(data1)

y_bar = colMeans(data1) - colMeans(data2)
y_bar

y_cov = cov(data1) + cov(data2)
y_cov

h_t = number_rows %*% t(y_bar) %*% solve(y_cov) %*% (y_bar)
h_t

f2 = ((number_rows - P)/ P) * (h_t/(number_rows-1))
f2

f_tab = qf(1-0.05, P, number_rows-P)
f_tab

if (f1 > f_tab) {
  print("We reject the null hypothesis at 5% level of Significance")
} else {
  print("We do not have enough evidence to reject the null hypothesis at 5% level of Significance")}
```
```{r}
data(iris)

setosa = iris[iris$Species == "setosa", 1:4]
versicolor = iris[iris$Species == "versicolor", 1:4]

head(setosa)

data1 = setosa
data2 = versicolor

number_rows = nrow(data1)

P = ncol(data1)

colMeans(data1)

y_bar = colMeans(data1) - colMeans(data2)
y_bar

y_cov = cov(data1) + cov(data2)
y_cov

h_t = number_rows %*% t(y_bar) %*% solve(y_cov) %*% (y_bar)
h_t

f2 = ((number_rows - P)/ P) * (h_t/(number_rows-1))
f2

f_tab = qf(1-0.05, P, number_rows-P)
f_tab

if (f1 > f_tab) {
  print("We reject the null hypothesis at 5% level of Significance")
} else {
  print("We do not have enough evidence to reject the null hypothesis at 5% level of Significance")}
```


