---
title: "PCA-PCR"
author: "Rupesh Pandey"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---


# STEPS TO PERFORM PRINCIPAL COMPONENT ANALYSIS :-
1. Obtain the eigen values and eigen vectors.
2. Obtain component loadings using the formula : g'sqrt(l)/sqrt(Sii)
3. obtain the percentage of variation of x explained by ith component is  λi/summation(λi)
    with the help of this percentage of variation, it explains how much component explains the variation
4. Anderson proposed the test procedure to test the equality of λi, using Bartlett test -- 
                the test statistics is : (n-(2p+1)/6)(p-r)ln(a0/g0)
                where, a0 = (lr+1 + lr+2 + ... + lp)/(p-r)
                     , g0 = (lr+1 . lr+2 . 
                     
   chi-square tabulated value :- chi-square(1/2)(p-r+2)(p-r-1)
   
# METHOD OF EXTRACTION OF PRINCIPAL COMPONENTS
-- Technique to transform the original set of variables into a set of linear combinations, so most of         variation in original dataset is explained by linear combinations. The linear combination so selected is    called principal components.


# PCA in Regression
-- PCA is used to orthogonalize X matrix. PCA transforms "X" into a new matrix "Z"
                          where, z = gamma'(x)
                                     -- matrix of eigen vectors and z is principal component.
                                replace X by Z then,
                                Y = Xbeta + E
                                  = Zbeta + E,
                                  = Gamma'Xbeta + E
                                  = ZGamma'Beta + E
                                  Intercept is 0
                                  
```{r}
var_cov = matrix(c(1.307,0.879,0.908,0.958,0.345,0.512,0.385,0,1.248,1.147,1.411,-0.089,0.564,0.172,0,0,1.317,1.480,0.058,0.573,0.259,0,0,0,2.022,-0.110,0.773,0.236,0,0,0,0,0.443,0.046,0.259,0,0,0,0,0,0.371,0.154,0,0,0,0,0,0,0.264),nrow = 7, ncol = 7, byrow = TRUE)
var_cov
var_c = var_cov + t(var_cov) - diag(diag(var_cov))
var_c

p = 7
n = 36

l = eigen(var_c)$value
g = eigen(var_c)$vector
l
g
g = round(g,4)

propvar = (l/sum(l))*100
propvar

cumprop = cumsum(propvar)
cumprop

pc_no = 1:length(l)
cbind(pc_no,l,propvar,cumprop)

rij = (t(g)*sqrt(l))/(sqrt(diag(var_c)))
rij = t(rij);rij
rij^2

for (r in 0:(p-1)) {
  a0 = sum(l[(r+1):p])/(p-r)
  g0 = prod(l[(r+1):p])^(1/(p-r))
  t = (n-((2*p+11)/6))*(p-r)*log(a0/g0)
  df = (p-r+2)*(p-r-1)/2

  pvalue = 1- pchisq(t,df)
  if(pvalue>= 0.05)
  {break}
}

cat("here accept H0 for r = ",r,"chisq =",t,"pvalue = ", pvalue)

plot(1:7,l,type = "b", col = "black")

```
```{r}
data = read.csv(file.choose())
data

var_covariance = cov(data[, 2:ncol(data)])
var_covariance
```

