---
title: "Extra functions of ggplot2"
author: "Sheetal Prasad"
date: "2025-09-30"
output: html_document
---

## ggcorrplot

The **`ggcorrplot`** package in R is used to visualize correlation matrices using **heatmaps with ggplot2**. It provides an intuitive and elegant way to explore the relationships between multiple variables in a dataset by representing correlations as heatmaps or circle plots. This package allows users to visualize not only the **strength** of the correlation, ranging from –1 to +1, but also the **direction** (positive or negative) through color gradients and shapes.

```{r}
# Load package
library(ggcorrplot)

```

Consider the mtcars dataset

```{r}
data(mtcars)

# Compute correlation matrix
corr <- cor(mtcars)

# Basic plot
ggcorrplot(corr)

```

Function involve in ggcorrplot package

-   `method` : Defines the shape of the plot, can be `"square"` or `"circle"`.

<!-- -->

-    `type` : Specifies which part of the matrix to display, options are `"full"`, `"lower"`, or `"upper"`.

-    `lab` : Logical, if `TRUE` displays the correlation coefficients on the plot.

<!-- -->

-    `lab_size` : Adjusts the text size of the displayed correlation coefficients.

<!-- -->

-    `colors` : Defines a custom color gradient for correlations.

<!-- -->

-    `hc.order` : Logical, if `TRUE` reorders the correlation matrix using hierarchical clustering

<!-- -->

-    `title` : Allows adding a custom title to the correlation plot.

<!-- -->

-    `ggtheme` : Specifies the ggplot2 theme to style the plot (e.g., `theme_minimal()`).

```{r}
# Customized correlation heatmap
ggcorrplot(corr,
           method = "circle",        # Shape: "square" or "circle"
           type = "lower",           # Show only lower triangle
           lab = TRUE,               # Display correlation coefficients
           lab_size = 3,             # Text size for labels
           colors = c("red", "white", "blue"), # Custom color gradient
           title = "Correlation Matrix of mtcars",
           ggtheme = ggplot2::theme_minimal())

```

## Heatmaply

The **`heatmaply`** package in R provides an efficient way to create interactive heatmaps that go beyond static visualizations. Unlike traditional heatmap functions, `heatmaply` builds upon the **plotly** library, enabling users to zoom, pan, hover over cells for details, and export plots as interactive HTML widgets.

```{r}
# Load library
library(heatmaply)
```

```{r}
data(mtcars)

# Create a basic heatmap
heatmaply(mtcars, 
          main = "Interactive Heatmap of mtcars Dataset")
```

```{r}
# Compute correlation matrix
corr_matrix <- cor(mtcars)

# Interactive heatmap of correlations
heatmaply(corr_matrix, 
          main = "Correlation Heatmap of mtcars",
          xlab = "Variables", 
          ylab = "Variables",
          k_col = 3,    # Cluster columns into 3 groups
          k_row = 3)    # Cluster rows into 3 groups

```

```{r}
heatmaply(mtcars, 
          scale = "column",                     # Scale data by column
          colors = colorRampPalette(RColorBrewer::brewer.pal(9, "Blues"))(100),
          dendrogram = "row",                   # Show dendrogram only for rows
          main = "Customized Heatmap with Dendrogram")

```

## **Key Functions and Options in `heatmaply`**

-    `heatmaply(x)` : Main function to generate an interactive heatmap.

-    `main` : Title of the heatmap.

-    `xlab`, `ylab` : Labels for x-axis and y-axis.

-   `scale` : Standardizes data by row, column, or none (`"row"`, `"column"`, `"none"`).

-    `colors` : Custom color palettes for heatmap cells.

-    `dendrogram` : Controls dendrogram display (`"both"`, `"row"`, `"column"`, or `"none"`).

-    `k_col`, `k_row` : Number of clusters for columns and rows.

-    `Rowv`, `Colv` : Logical, determines whether rows/columns are reordered.

-    `hide_colorbar` : Hides or displays the color legend.

-    `file` : Saves the heatmap to an external HTML file for sharing.

## Reference

1\. <https://librarycarpentry.github.io/lc-r/instructor/04-data-viz-ggplot.html#>:\~:text=The%20data%20argument%20is%20used,barplot%2C%20scatter%20plot%2C%20boxplot.

2.  <https://www.geeksforgeeks.org/r-language/data-visualization-with-r-and-ggplot2/>

3\. <https://clauswilke.com/dataviz/aesthetic-mapping.html>
